---
title: "Supplementary Analysis: Excitatory Neuron Classification in Postnatal PsychAD Cells"
subtitle: "Root-cause evaluation of low EN fraction in scANVI-integrated infant transcriptomes"
author: "R.A. Dear"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
    fig-width: 8
    fig-height: 4.5
    number-sections: true
engine: knitr
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(yaml)
library(patchwork)
library(scales)

# Read config path from environment (set by render_notebook.sh via NOTEBOOK_PARAMS)
params_file <- Sys.getenv("NOTEBOOK_PARAMS", unset = NA_character_)
if (!is.na(params_file) && file.exists(params_file)) {
  cfg  <- yaml::read_yaml(params_file)
  BASE <- cfg$BASE
} else {
  BASE <- "/home/rajd2/rds/hpc-work/snRNAseq_2026/scripts/relabel_comparison"
}

# Colour scheme — consistent across all figures
GROUP_COLORS <- c(
  "PsychAD <1y"   = "#C0392B",
  "PsychAD 1–5y"  = "#E07B5A",
  "Wang <1y"      = "#2980B9",
  "Vel <1y"       = "#27AE60",
  "PsychAD"       = "#C0392B",
  "Wang"          = "#2980B9",
  "Vel"           = "#27AE60"
)

# Shared ggplot theme
theme_pub <- function(...) {
  theme_classic(base_size = 11) +
    theme(
      strip.background  = element_blank(),
      strip.text        = element_text(face = "bold", size = 10),
      axis.title        = element_text(size = 10),
      axis.text         = element_text(size = 9),
      legend.title      = element_blank(),
      legend.text       = element_text(size = 9),
      legend.key.size   = unit(0.45, "cm"),
      plot.title        = element_text(face = "bold", size = 11),
      plot.subtitle     = element_text(size = 9, colour = "grey40"),
      ...
    )
}
```

## Background {.unnumbered}

In the integrated snRNAseq dataset combining Wang (2022), Velmeshev (2022; hereafter Vel), and
PsychAD (Mathys 2023), scANVI-based label transfer assigns approximately 5% of PsychAD infant
(<1 year) cells to the Excitatory (EN) class. By contrast, Wang and Vel cells from the same
developmental window carry EN fractions of ~25% and ~30%, respectively. Given that the prefrontal
cortex is characterised by substantial excitatory neuron populations at birth
(*in utero* EN migration is largely complete by mid-gestation), a 5-fold deficit raises the question of
whether the low EN fraction reflects (i) genuine biological differences in the PsychAD cohort,
(ii) a technical or normalisation artefact, or (iii) a misclassification error arising from the
integration model.

This supplementary analysis applies a series of orthogonal marker-based diagnostics to the raw,
unnormalised count matrices of each source. The principal conclusion is that the EN fraction in
PsychAD infants is genuinely low relative to Wang/Vel — partly reflecting a biological signature
of early postnatal EN immaturity — but is further suppressed by a label-transfer artefact caused
by the dominance of 453,000 adult PsychAD cells in the scANVI supervision signal.

---

## Hypothesis H9 Rejected: Library Depth and HVG Selection Are Not Confounders

### Per-cell library depth

A straightforward confound would arise if PsychAD cells have systematically lower UMI counts than
Wang/Vel, causing EN markers to fall below detection thresholds under CPM normalisation. We
examined total UMI depth for the matched <1 year subsets from each source.

```{r fig-depth, fig.cap="**Figure S1.** Per-cell library depth (total UMI counts) for the matched infant (<1 year) subsets. Bars show the inter-quartile range (P25–P75); the point marks the mean; the horizontal line within the bar marks the median. PsychAD cells carry approximately twice the UMI depth of Wang and Vel cells, ruling out sparse-count normalisation as an explanation for the observed EN deficit."}
qc <- read_csv(file.path(BASE, "hvg_intersection_v2", "per_cell_qc.csv"), show_col_types = FALSE) |>
  mutate(
    group_label = recode(group,
      "PSYCHAD" = "PsychAD",
      "WANG"    = "Wang",
      "VEL"     = "Vel"
    ),
    group_label = factor(group_label, levels = c("PsychAD", "Wang", "Vel"))
  )

p_depth <- ggplot(qc, aes(x = group_label, fill = group_label)) +
  geom_col(aes(y = total_counts_median), colour = "white", width = 0.55) +
  geom_errorbar(
    aes(ymin = total_counts_P25, ymax = total_counts_P75),
    width = 0.22, linewidth = 0.6, colour = "grey30"
  ) +
  geom_point(aes(y = total_counts_mean), shape = 21, fill = "white",
             colour = "grey20", size = 2.5, stroke = 0.8) +
  scale_fill_manual(values = GROUP_COLORS, guide = "none") +
  scale_y_continuous(
    labels = label_comma(),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    x     = NULL,
    y     = "UMI count per cell",
    title = "Library depth: PsychAD is ~2× deeper than Wang/Vel"
  ) +
  theme_pub()

p_depth
```

PsychAD infant cells have a median of **8,636 UMI/cell** (mean 12,587), compared with 5,486 (mean
6,508) for Wang and 4,319 (mean 5,560) for Vel. Because CPM normalisation scales expression
*inversely* with library size, deeper PsychAD libraries would, if anything, *reduce* normalised
marker counts per unit of genuine expression — making this hypothesis untenable in the direction
that would inflate IN relative to EN.

### EN:IN ratio is invariant across gene-set sizes

A second technical concern is that HVG selection could preferentially capture IN markers while
excluding EN markers when Wang/Vel gene space is used as the reference. We evaluated the EN:IN
marker ratio across HVG intersection sizes ranging from 423 to 15,000 genes.

```{r fig-hvg, fig.cap="**Figure S2.** EN-to-IN marker mean CPM ratio across five HVG intersection sizes (cross-dataset intersections of Seurat PearsonResiduals HVGs). Each source group is independently normalised before computing the ratio. At ≥1,374 genes (2,500 nominal HVGs), the ratio stabilises: PsychAD ≈ 0.16, Wang ≈ 1.1–1.3, Vel ≈ 3.0–4.1. The 7–25-fold deficit is consistent across gene-set sizes, rejecting a gene-selection explanation."}
hvg <- read_csv(file.path(BASE, "hvg_intersection_v2", "markers_by_hvg_size.csv"),
                show_col_types = FALSE) |>
  filter(intersection_size >= 1000) |>   # ≥1000 genes: all EN markers captured
  mutate(
    group_label = recode(group,
      "PSYCHAD" = "PsychAD",
      "WANG"    = "Wang",
      "VEL"     = "Vel"
    ),
    group_label = factor(group_label, levels = c("PsychAD", "Wang", "Vel"))
  )

p_hvg <- ggplot(hvg, aes(x = intersection_size, y = en_over_in,
                          colour = group_label, group = group_label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3, stroke = 0.5, fill = "white", shape = 21) +
  scale_colour_manual(values = GROUP_COLORS) +
  scale_x_continuous(labels = label_comma(), breaks = c(1000, 3000, 5000, 8000, 15000)) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.08))) +
  annotate("text", x = 14500, y = 4.25, label = "Vel",
           colour = GROUP_COLORS["Vel"], size = 3.5, fontface = "bold") +
  annotate("text", x = 14500, y = 1.35, label = "Wang",
           colour = GROUP_COLORS["Wang"], size = 3.5, fontface = "bold") +
  annotate("text", x = 14500, y = 0.08, label = "PsychAD",
           colour = GROUP_COLORS["PsychAD"], size = 3.5, fontface = "bold") +
  labs(
    x     = "HVG intersection size (genes)",
    y     = "EN / IN marker mean CPM",
    title = "EN:IN ratio is stable across gene-set sizes",
    subtitle = "PsychAD ≈ 0.16, Wang ≈ 1.2, Vel ≈ 3.0 — independent of HVG choice"
  ) +
  theme_pub() +
  theme(legend.position = "none")

p_hvg
```

Across all intersection sizes ≥1,374 genes (at which point both EN and IN marker sets are fully
represented), the EN:IN ratio is stable within each source. The ratio for PsychAD plateaus at
0.16, while Wang and Vel maintain 1.1–1.3 and 3.0–4.1, respectively. This 7–25-fold differential
does not narrow as more genes are included, definitively rejecting gene-selection bias as an
explanation.

---

## Raw Marker Expression Confirms a Genuine EN Deficit in PsychAD Infants

Having excluded technical confounders, we examined raw count-derived marker expression using
CPM values computed directly from each source's un-integrated count matrix.

```{r fig-cpm, fig.height=5.5, fig.cap="**Figure S3.** Mean CPM of established excitatory (EN) and inhibitory (IN) neuron markers computed directly from the raw unnormalised count matrix of each source. Groups shown are the matched postnatal subsets (<1 year and 1–5 years for PsychAD). EN marker expression is depressed 4–11× in PsychAD <1y relative to Wang/Vel despite comparable or greater library depth (Fig. S1). IN marker expression is elevated in PsychAD relative to the other sources. Prenatal Wang and Vel groups are shown as reference for the expected mature EN state."}
cpm_raw <- read_csv(file.path(BASE, "no_age_run", "marker_means_raw_cpm.csv"),
                    show_col_types = FALSE) |>
  filter(group %in% c("PSYCHAD_under1y", "PSYCHAD_1_5y",
                       "WANG_under1y",    "VEL_V3_under1y")) |>
  filter(marker %in% c("SATB2", "SLC17A7", "NEUROD2", "RBFOX3",
                        "GAD1",  "GAD2",    "LHX6")) |>
  mutate(
    group_label = recode(group,
      "PSYCHAD_under1y" = "PsychAD <1y",
      "PSYCHAD_1_5y"    = "PsychAD 1–5y",
      "WANG_under1y"    = "Wang <1y",
      "VEL_V3_under1y"  = "Vel <1y"
    ),
    group_label = factor(group_label,
                         levels = c("PsychAD <1y", "PsychAD 1–5y",
                                    "Wang <1y", "Vel <1y")),
    marker_type_label = if_else(marker_type == "EN",
                                "Excitatory (EN) markers",
                                "Inhibitory (IN) markers"),
    marker = factor(marker, levels = c("SATB2", "SLC17A7", "NEUROD2", "RBFOX3",
                                        "GAD1",  "GAD2",    "LHX6"))
  )

p_cpm <- ggplot(cpm_raw, aes(x = marker, y = mean_norm, fill = group_label)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7, colour = "white") +
  scale_fill_manual(values = GROUP_COLORS) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  facet_wrap(~marker_type_label, scales = "free") +
  labs(
    x     = "Marker gene",
    y     = "Mean CPM",
    title = "EN markers are 4–11× lower in PsychAD <1y; IN markers are elevated"
  ) +
  theme_pub() +
  theme(
    legend.position = "bottom",
    axis.text.x     = element_text(face = "italic", angle = 30, hjust = 1)
  )

p_cpm
```

SATB2 — a high-specificity excitatory neuron transcription factor widely used to quantify EN
content in developing cortex — shows a mean CPM of **0.73** in PsychAD <1y versus **3.48** in
Wang and **8.42** in Vel, a 4.8–11.5-fold deficit. SLC17A7 (vGLUT1) and NEUROD2 show comparable
or larger deficits. Conversely, GAD1 (0.237 CPM in Wang, 0.117 in Vel, **2.37** in PsychAD <1y)
and GAD2 are substantially elevated in PsychAD, producing an inverted EN:IN balance at the
transcriptomic level.

This pattern is consistent with a genuine preponderance of cortical interneurons at the expense of
pyramidal cells in the PsychAD DLPFC sample. However, the magnitude of the deficit relative to
Wang and Vel — sourced from the same developmental window — suggests additional sources of
variance that we examine below.

---

## Per-Cell Marker Detectability Confirms the Composition Shift

The CPM analysis above reflects mean expression per cell summed across the population; high
expression in a minority of cells can inflate the mean. We therefore examined the fraction of
cells with detectable expression (UMI ≥ 1) for each marker independently.

```{r fig-frac, fig.cap="**Figure S4.** Fraction of cells with detectable expression (≥1 UMI) of EN and IN marker genes, computed from raw counts. Each dot represents one marker gene; the bar shows the mean across markers within each class. Dashed lines connect the EN and IN mean for each group, highlighting the EN/IN contrast. Across all EN markers, PsychAD <1y shows ~12% detectability versus ~48% (Wang) and ~62% (Vel). IN detectability is substantially higher in PsychAD (<1y: 44%) than in the reference cohorts (~20%)."}
d12 <- read_csv(file.path(BASE, "d11_d12_d13", "d12_frac_nonzero.csv"),
                show_col_types = FALSE) |>
  mutate(
    group_label = recode(group,
      "PSYCHAD_under1y" = "PsychAD <1y",
      "PSYCHAD_1_5y"    = "PsychAD 1–5y",
      "WANG_under1y"    = "Wang <1y",
      "VEL_V3_under1y"  = "Vel <1y"
    ),
    group_label = factor(group_label,
                         levels = c("PsychAD <1y", "PsychAD 1–5y",
                                    "Wang <1y",    "Vel <1y"))
  ) |>
  select(group_label,
         EN_SATB2    = EN_SATB2_frac1,
         EN_SLC17A7  = EN_SLC17A7_frac1,
         EN_NEUROD2  = EN_NEUROD2_frac1,
         EN_RBFOX3   = EN_RBFOX3_frac1,
         EN_TBR1     = EN_TBR1_frac1,
         IN_GAD1     = IN_GAD1_frac1,
         IN_GAD2     = IN_GAD2_frac1,
         IN_LHX6     = IN_LHX6_frac1,
         IN_SLC32A1  = IN_SLC32A1_frac1) |>
  pivot_longer(-group_label, names_to = c("type", "marker"),
               names_sep = "_(?=[A-Z])", values_to = "frac_nonzero") |>
  mutate(
    marker_type_label = if_else(type == "EN",
                                "Excitatory (EN) markers",
                                "Inhibitory (IN) markers")
  )

# Means for dashed connectors
d12_means <- d12 |>
  group_by(group_label, marker_type_label) |>
  summarise(mean_frac = mean(frac_nonzero), .groups = "drop") |>
  pivot_wider(names_from = marker_type_label, values_from = mean_frac)

d12_long_means <- d12 |>
  group_by(group_label, marker_type_label) |>
  summarise(mean_frac = mean(frac_nonzero), .groups = "drop")

p_frac <- ggplot(d12, aes(x = group_label, y = frac_nonzero, colour = group_label)) +
  geom_jitter(width = 0.12, size = 2.5, alpha = 0.8, stroke = 0.3) +
  stat_summary(fun = mean, geom = "crossbar",
               width = 0.55, linewidth = 0.6, fatten = 1.5,
               colour = "grey20") +
  scale_colour_manual(values = GROUP_COLORS, guide = "none") +
  scale_y_continuous(labels = label_percent(accuracy = 1),
                     limits = c(0, 0.85),
                     expand = expansion(mult = c(0, 0.05))) +
  facet_wrap(~marker_type_label) +
  labs(
    x     = NULL,
    y     = "Fraction of cells (≥1 UMI)",
    title = "EN detectability 4–5× lower in PsychAD; IN detectability elevated"
  ) +
  theme_pub() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p_frac
```

The frac_nonzero metric is unaffected by CPM normalisation or library size. Across five EN
markers, PsychAD <1y shows a mean detectability of **12%** (range 3–20%), compared with **24–48%**
for Wang and **13–62%** for Vel. The IN equivalent for PsychAD <1y is **44%**, substantially
above the reference cohorts (20% and 19%, respectively). The EN/IN detectability ratio for
PsychAD <1y is 0.26, versus 1.21 for Wang and 1.65 for Vel — a consistent 5–6-fold difference
that cannot be attributed to normalisation choices.

Notably, the 1–5 year PsychAD group shows intermediate EN detectability (15%) and reduced IN
detectability (28%) relative to the <1y group, suggesting a partial developmental trajectory —
albeit one that does not reach the Wang/Vel level even by age 5.

---

## Within-Cohort Donor Analysis Shows No Age Gradient

A potential confound within the PsychAD <1y group is age heterogeneity: if the youngest donors
(1 month) have genuinely lower EN fractions than donors at 4–6 months, the bulk aggregate could
mask a biologically coherent developmental signal.

```{r fig-donor, fig.width=7, fig.height=4, fig.cap="**Figure S5.** Per-donor SATB2 detectability (fraction of cells ≥1 UMI; *n* = 11 donors, 1–6 months) against donor age. Point size encodes the number of cells. Horizontal dashed lines mark the Wang (blue) and Vel (green) group-level SATB2 detectability for the matched <1y subset. No consistent age gradient is observed: the two 6-month donors are the most discordant pair in the cohort (0.115 vs. 0.294), and the 1-month donor (Donor_122) lies at the cohort median (0.219). SATB2 retention — the fraction of SATB2⁺ cells (≥1 UMI) that also express ≥3 UMI — is uniformly low across donors (mean 0.31; Wang 0.77; Vel 0.75), indicating a systematic early postnatal feature rather than a dropout artefact."}
d11 <- read_csv(file.path(BASE, "d11_d12_d13", "d11_per_donor.csv"),
                show_col_types = FALSE)

# Wang and Vel SATB2_frac1 reference lines from d12
wang_satb2 <- d12 |>
  filter(group_label == "Wang <1y", marker == "SATB2") |>
  pull(frac_nonzero)
vel_satb2 <- d12 |>
  filter(group_label == "Vel <1y",  marker == "SATB2") |>
  pull(frac_nonzero)

p_donor <- ggplot(d11, aes(x = age_months, y = SATB2_frac1)) +
  geom_hline(yintercept = wang_satb2,
             colour = GROUP_COLORS["Wang <1y"], linetype = "dashed", linewidth = 0.7) +
  geom_hline(yintercept = vel_satb2,
             colour = GROUP_COLORS["Vel <1y"],  linetype = "dashed", linewidth = 0.7) +
  annotate("text", x = 6.3, y = wang_satb2 + 0.015,
           label = paste0("Wang <1y (", percent(wang_satb2, accuracy = 0.1), ")"),
           colour = GROUP_COLORS["Wang <1y"], size = 3, hjust = 1) +
  annotate("text", x = 6.3, y = vel_satb2 + 0.015,
           label = paste0("Vel <1y (", percent(vel_satb2, accuracy = 0.1), ")"),
           colour = GROUP_COLORS["Vel <1y"], size = 3, hjust = 1) +
  geom_point(aes(size = n_cells),
             colour = GROUP_COLORS["PsychAD <1y"], alpha = 0.85, stroke = 0.3) +
  geom_smooth(method = "lm", formula = y ~ x,
              colour = "grey40", fill = "grey85", linewidth = 0.7, se = TRUE) +
  scale_x_continuous(breaks = 1:6, limits = c(0.7, 7)) +
  scale_y_continuous(labels = label_percent(accuracy = 1),
                     limits = c(0, 0.75),
                     expand = expansion(mult = c(0.02, 0.05))) +
  scale_size_area(max_size = 9, name = "Cells per donor") +
  labs(
    x     = "Donor age (months)",
    y     = "SATB2 fraction of cells (≥1 UMI)",
    title = "No consistent age gradient within PsychAD <1y",
    subtitle = "All donors are ~4–11× below Wang/Vel; 6-month donors are the most discordant pair"
  ) +
  theme_pub() +
  theme(legend.position = "right")

p_donor
```

The linear trend across 1–6 months is flat and non-significant; the two 6-month donors (Donor_202:
11.5%; Donor_503: 29.4%) bracket the entire cohort distribution. SATB2 retention (≥3 UMI / ≥1 UMI,
measuring per-cell expression intensity among SATB2-expressing cells) has a donor mean of
**0.31 ± 0.08** in PsychAD <1y versus **0.77** in Wang and **0.75** in Vel — a deficit that is
consistent across all 11 donors and not explained by any obvious technical covariate (Table S1
shows per-donor metadata).

```{r tbl-donor}
knitr::kable(
  d11 |>
    select(Donor = individual,
           `Age (mo.)` = age_months,
           `Stage` = development_stage,
           `N cells` = n_cells,
           `UMI/cell` = total_UMI_mean,
           `SATB2 frac1` = SATB2_frac1,
           `SATB2 ret.` = SATB2_retention,
           `GAD1 frac1` = GAD1_frac1) |>
    mutate(across(where(is.double), ~round(., 3))),
  caption = "**Table S1.** Per-donor summary for PsychAD <1y group (sorted by age).
  SATB2 ret. = fraction of SATB2-positive cells with ≥3 UMI (expression intensity proxy).
  Donor_920 has only 392 cells and should be interpreted cautiously.",
  align = c("l", "r", "l", "r", "r", "r", "r", "r")
)
```

---

## Ambient RNA Contamination Does Not Explain Elevated IN Signal

High IN marker CPM in PsychAD cells could arise not from genuine interneuron content but from
ambient RNA contamination if the droplet suspension is enriched for nuclear fragments expressing
GAD1/GAD2. We tested this by computing IN marker levels in cells classified as *non-neuronal*
by the source-native PsychAD annotation (HBCC cohort `class` column: Astro, Oligo, OPC, Mural,
Endo, Immune) — cells that should not express interneuron markers in the absence of contamination.

```{r fig-ambient, fig.cap="**Figure S6.** IN marker expression in non-neuronal cells, classified by source-native labels (HBCC `class` for PsychAD; scANVI `cell_type_aligned` for Wang/Vel). CPM is computed from raw counts for cells not assigned to EN or IN classes. If elevated GAD1/GAD2 in PsychAD non-neuronal cells were driven by ambient RNA, the PsychAD bars would substantially exceed Wang/Vel. The observed pattern shows PsychAD intermediate between Wang and Vel for GAD1, and comparable for GAD2 and SLC32A1. This argues against ambient contamination as the explanation for elevated IN signal in PsychAD infants."}
d13b <- read_csv(file.path(BASE, "d11_d12_d13", "d13b_ambient_native_labels.csv"),
                 show_col_types = FALSE) |>
  mutate(
    group_label = recode(group,
      "PSYCHAD_under1y" = "PsychAD <1y",
      "PSYCHAD_1_5y"    = "PsychAD 1–5y",
      "WANG_under1y"    = "Wang <1y",
      "VEL_V3_under1y"  = "Vel <1y"
    ),
    group_label = factor(group_label,
                         levels = c("PsychAD <1y", "PsychAD 1–5y",
                                    "Wang <1y", "Vel <1y"))
  ) |>
  select(group_label, label_source, n_nonneuronal,
         GAD1 = GAD1_mean_cpm, GAD2 = GAD2_mean_cpm, SLC32A1 = SLC32A1_mean_cpm) |>
  pivot_longer(c(GAD1, GAD2, SLC32A1), names_to = "marker", values_to = "mean_cpm") |>
  mutate(marker = factor(marker, levels = c("GAD1", "GAD2", "SLC32A1")))

p_ambient <- ggplot(d13b, aes(x = marker, y = mean_cpm, fill = group_label)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7, colour = "white") +
  scale_fill_manual(values = GROUP_COLORS) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(
    x     = "Marker gene",
    y     = "Mean CPM (non-neuronal cells only)",
    title = "IN marker background is comparable across sources",
    subtitle = "PsychAD non-neuronal cells classified by HBCC native class labels (not scANVI)"
  ) +
  theme_pub() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(face = "italic"))

p_ambient
```

```{r tbl-ambient}
d13b_wide <- read_csv(file.path(BASE, "d11_d12_d13", "d13b_ambient_native_labels.csv"),
                      show_col_types = FALSE) |>
  mutate(
    group_label = recode(group,
      "PSYCHAD_under1y" = "PsychAD <1y",
      "PSYCHAD_1_5y"    = "PsychAD 1–5y",
      "WANG_under1y"    = "Wang <1y",
      "VEL_V3_under1y"  = "Vel <1y"
    )
  ) |>
  select(`Group` = group_label,
         `Label source` = label_source,
         `Non-neuronal` = n_nonneuronal,
         `Neuronal` = n_neuronal,
         `EN CPM` = EN_mean_cpm,
         `IN CPM` = IN_mean_cpm,
         `IN/EN` = IN_over_EN_nonneuronal,
         `GAD1 CPM` = GAD1_mean_cpm) |>
  mutate(across(where(is.double), ~round(., 3)))

knitr::kable(d13b_wide,
  caption = "**Table S2.** IN/EN marker CPM in non-neuronal cells by source. Label source indicates
  which annotation was used to identify non-neuronal cells. The IN/EN ratio in PsychAD non-neuronal
  cells is elevated relative to Wang/Vel but is consistent with the Vel prenatal group, which also
  uses FANS nuclear isolation and has well-characterised ambient contamination levels."
)
```

GAD1 CPM in PsychAD <1y non-neuronal cells (0.629) is intermediate between Wang (0.464) and Vel
(0.833). GAD2 and SLC32A1 are uniformly low across all sources. Critically, the classification of
PsychAD non-neuronal cells by the HBCC native annotation (`class` column) and by scANVI-derived
`cell_type_aligned` differs by only 7 cells (of 15,037), confirming that the ambient RNA test is
not circular with respect to the scANVI model under investigation.

Both PsychAD and Vel use FANS (fluorescence-activated nuclear sorting) for single-nucleus isolation,
which substantially reduces cytoplasmic ambient RNA. The comparable GAD1 background across these
FANS-sorted datasets argues against contamination as the primary explanation for elevated IN signal
in PsychAD infants.

---

## Summary of Marker-Based Diagnostics

The diagnostic analyses above support the following decomposition of the low EN fraction in
PsychAD <1y cells:

| Component | Evidence | Magnitude |
|-----------|----------|-----------|
| **Genuine biology** | SATB2 frac_nonzero 12% vs 48–62%; stable across donors; no library-size confound | ~4–5× EN deficit relative to Wang/Vel |
| **Annotation bias (scANVI)** | 453k adult PsychAD cells dominate supervision signal; scANVI assigns ~5% EN vs raw marker floor of ~12–20% | ~2–4× additional suppression |
| **Ambient RNA** | Non-neuronal GAD1 in PsychAD comparable to Wang/Vel | Not a primary driver |

: **Table S3.** Decomposition of the PsychAD <1y EN fraction deficit. {tbl-colwidths="[25,55,20]"}

The raw marker analysis establishes a biological floor of approximately 12–20% EN cells in
PsychAD <1y (SATB2 frac_nonzero 19.5%, mean CPM-inferred proportion ~12%). These analyses
motivated removing adult PsychAD supervision from scANVI to obtain an unbiased estimate of the
true EN fraction. The scANVI cross-assignment experiments described in the following sections
provide a definitive answer.

---

## scANVI Cross-Assignment Experiments

Two complementary scANVI experiments were run on a focused dataset: Velmeshev (Vel) and PsychAD
cells only, age ≤5 years, 20,000 cells per source per age bin (<1y, 1–5y), no Wang cells.
The experiments swap the direction of supervision to establish the ground truth for both
the Vel "Interneurons" label mystery and the PsychAD <1y IN-dominated composition.

### Context: The Vel "Interneurons" Annotation Problem

Velmeshev et al. (2022) annotated their postnatal cortical dataset using a label vocabulary
that differs substantially from the adult PsychAD (Mathys 2023) subclass nomenclature.
Of the 20,000 Vel cells in the <1y age bin used in these experiments, **57.4% carry the label
"Interneurons"** — a term that in adult cortex unambiguously refers to GABAergic inhibitory
neurons. A further 14.3% are labelled "INT" (a second Vel IN-like category). The question
is whether these cells are genuinely inhibitory, or whether they represent a class of immature
excitatory progenitors that Velmeshev grouped under a developmentally meaningful but
nomenclaturally ambiguous label.

```{r vel-native-labels, fig.height=4, fig.cap="**Figure S7.** Velmeshev native (paper) cell type labels for the <1y age bin used in the cross-assignment experiments (n = 20,000 cells). 'Interneurons' is the single largest class (57.4%), followed by 'INT' (14.3%), both potentially inhibitory by name. Whether these cells are transcriptionally EN or IN is the subject of Experiments 1 and 2."}
vel_native <- tribble(
  ~label,                    ~pct,  ~broad_class,
  "Interneurons",            57.4,  "Ambiguous",
  "INT",                     14.3,  "Ambiguous",
  "Glial_progenitors",        7.2,  "Glia",
  "Progenitors",              6.4,  "Glia",
  "Protoplasmic_astrocytes",  4.8,  "Glia",
  "OPC",                      3.0,  "Glia",
  "Fibrous_astrocytes",       1.9,  "Glia",
  "Microglia",                1.0,  "Glia",
  "SST_RELN",                 0.6,  "IN",
  "SST",                      0.5,  "IN",
  "CALB2",                    0.4,  "IN",
  "PV",                       0.4,  "IN",
  "L4",                       0.4,  "EN",
  "CCK",                      0.3,  "IN",
  "VIP",                      0.3,  "IN"
) |>
  mutate(
    label = factor(label, levels = rev(label)),
    broad_class = factor(broad_class, levels = c("Ambiguous","EN","IN","Glia"))
  )

CLASS_COLS <- c("Ambiguous" = "#8E44AD", "EN" = "#2980B9", "IN" = "#E67E22", "Glia" = "#7F8C8D")

ggplot(vel_native, aes(x = label, y = pct, fill = broad_class)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = CLASS_COLS) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08)),
                     labels = label_percent(scale = 1)) +
  labs(x = NULL, y = "% of Vel <1y cells",
       title = "Vel <1y native labels: 'Interneurons' dominates",
       subtitle = "Purple = ambiguous label; cross-assignment resolves EN vs IN identity",
       fill = "Broad class") +
  theme_pub()
```

### Experiment 1: Vel Native Labels → Classify PsychAD Cells

**Design:** scANVI was trained using Vel native (paper) labels as supervision
(`use_raw_labels: true`) and all PsychAD cells as Unknown. scANVI must therefore assign
each PsychAD cell to the closest Vel native type — including "Interneurons", "INT", "L4",
"SST", "PV", "VIP", etc.

**Key question:** Do PsychAD <1y cells cluster with Vel "Interneurons" (which would suggest
they are EN-like), or with canonical IN types (SST/PV/VIP)?

```{r exp1-psychad-result, fig.height=5, fig.cap="**Figure S8.** Classification of PsychAD <1y cells (n = 20,000) under Vel native label supervision (Experiment 1). Each PsychAD cell is assigned the nearest Vel native type by scANVI. Only 2.4% land on 'Interneurons'; the majority land on OPC (27.6%), Microglia (14.4%), and canonical inhibitory types — SST (7.8%), PV (7.3%), VIP (5.8%), INT (5.5%), CALB2 (5.3%), SST_RELN (5.0%). This result demonstrates that PsychAD <1y cells are transcriptionally closer to cardinal interneurons than to Vel 'Interneurons'."}
exp1_psych <- tribble(
  ~label,                    ~pct,  ~broad_class,
  "OPC",                     27.6,  "Glia",
  "Microglia",               14.4,  "Glia",
  "SST",                      7.8,  "IN",
  "PV",                       7.3,  "IN",
  "VIP",                      5.8,  "IN",
  "INT",                      5.5,  "Ambiguous",
  "CALB2",                    5.3,  "IN",
  "SST_RELN",                 5.0,  "IN",
  "Glial_progenitors",        3.6,  "Glia",
  "CCK",                      3.3,  "IN",
  "L6",                       2.7,  "EN",
  "SV2C",                     2.6,  "IN",
  "Interneurons",             2.4,  "Ambiguous",
  "RELN",                     2.0,  "IN",
  "NOS",                      1.9,  "IN"
) |>
  mutate(
    label = factor(label, levels = rev(label)),
    broad_class = factor(broad_class, levels = c("Ambiguous","EN","IN","Glia"))
  )

ggplot(exp1_psych, aes(x = label, y = pct, fill = broad_class)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = CLASS_COLS) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08)),
                     labels = label_percent(scale = 1)) +
  labs(x = NULL, y = "% of PsychAD <1y cells assigned to Vel class",
       title = "PsychAD <1y → canonical IN types, not 'Interneurons'",
       subtitle = "Only 2.4% of PsychAD <1y land on Vel 'Interneurons'; 46.7% land on IN types",
       fill = "Broad class") +
  theme_pub()
```

**Interpretation:** PsychAD <1y cells are classified as genuine canonical inhibitory neurons
(SST, PV, VIP) at high frequency. The 2.4% that land on "Interneurons" represents a small minority.
This rules out the hypothesis that PsychAD "immature neurons" are simply the same cells as
Vel "Interneurons" expressed in a different latent space due to batch effects.
The PsychAD <1y IN-enrichment is real.

### Experiment 2: PsychAD Native Labels → Classify Vel Cells

**Design:** scANVI was trained using PsychAD native HBCC subclass labels as supervision
(`use_raw_labels: true`) and all Vel cells as Unknown. Each Vel cell is assigned the nearest
PsychAD subclass (EN_L2_3_IT, IN_SST, IN_PVALB, OPC, etc.). The scVI model from Experiment 1
was reused (symlink); only the scANVI label assignment changes.

**Key question:** When placed in PsychAD label space, do Vel "Interneurons" land on EN or IN
subtypes?

```{r exp2-vel-interneurons, fig.height=4.5, fig.cap="**Figure S9.** Top PsychAD subclass assignments for Vel 'Interneurons' (n = 16,411 cells) under PsychAD supervision (Experiment 2). 96.1% are classified as EN subtypes, with EN_L2_3_IT (upper-layer IT neurons) dominating at 45.6% (median confidence 0.94). Only 1.2% land on IN subtypes. This definitively identifies Vel 'Interneurons' as immature upper-layer excitatory neurons, not inhibitory neurons."}
vel_interneurons_exp2 <- tribble(
  ~label,          ~pct,   ~med_conf, ~broad_class,
  "EN_L2_3_IT",    45.6,   0.94,      "EN",
  "EN_L6_IT_1",    14.1,   0.87,      "EN",
  "EN_L6_CT",      12.2,   0.65,      "EN",
  "EN_L6B",        11.9,   0.84,      "EN",
  "EN_L3_5_IT_2",   5.6,   0.80,      "EN",
  "EN_L3_5_IT_3",   4.4,   0.99,      "EN",
  "OPC",            1.1,   0.54,      "Glia",
  "EN_L3_5_IT_1",   1.1,   0.79,      "EN",
  "Endo",           0.7,   0.33,      "Glia",
  "EN_L5_6_NP",     0.6,   0.90,      "EN",
  "EN_L5_ET",       0.5,   0.85,      "EN",
  "IN_PVALB",       0.4,   0.56,      "IN",
  "IN_SST",         0.4,   0.60,      "IN"
) |>
  mutate(
    label = factor(label, levels = rev(label)),
    broad_class = factor(broad_class, levels = c("EN","IN","Glia"))
  )

EN_IN_COLS <- c("EN" = "#2980B9", "IN" = "#E67E22", "Glia" = "#7F8C8D")

ggplot(vel_interneurons_exp2, aes(x = label, y = pct, fill = broad_class)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = EN_IN_COLS) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08)),
                     labels = label_percent(scale = 1)) +
  labs(x = NULL, y = "% of Vel 'Interneurons' (n = 16,411)",
       title = "Vel 'Interneurons' → 96.1% EN in PsychAD label space",
       subtitle = "Predominantly EN_L2_3_IT (upper-layer IT neurons); mean confidence 0.77",
       fill = NULL) +
  theme_pub()
```

```{r exp2-vel-int, fig.height=4, fig.cap="**Figure S10.** Top PsychAD subclass assignments for Vel 'INT' (n = 2,922 cells). 88.2% land on canonical inhibitory subtypes (PV, SST, VIP, LAMP5), confirming that Vel 'INT' represents genuine cortical interneurons. The sharp contrast with 'Interneurons' (96.1% EN) validates the EN/IN distinction within the Vel vocabulary."}
vel_int_exp2 <- tribble(
  ~label,             ~pct,  ~broad_class,
  "IN_PVALB",         22.6,  "IN",
  "IN_SST",           17.0,  "IN",
  "IN_VIP",           16.4,  "IN",
  "IN_LAMP5_LHX6",    14.0,  "IN",
  "IN_LAMP5_RELN",    12.0,  "IN",
  "IN_ADARB2",         5.8,  "IN",
  "Endo",              5.5,  "Glia",
  "EN_L6_CT",          1.6,  "EN",
  "OPC",               1.6,  "Glia",
  "EN_L6B",            1.2,  "EN",
  "IN_PVALB_CHC",      0.3,  "IN"
) |>
  mutate(
    label = factor(label, levels = rev(label)),
    broad_class = factor(broad_class, levels = c("EN","IN","Glia"))
  )

ggplot(vel_int_exp2, aes(x = label, y = pct, fill = broad_class)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = EN_IN_COLS) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08)),
                     labels = label_percent(scale = 1)) +
  labs(x = NULL, y = "% of Vel 'INT' (n = 2,922)",
       title = "Vel 'INT' → 88.2% genuine IN (PV/SST/VIP/LAMP5)",
       subtitle = "Contrast with 'Interneurons': INT is truly inhibitory, Interneurons are excitatory progenitors",
       fill = NULL) +
  theme_pub()
```

### Vel Native Label × PsychAD Subclass Cross-Tab

```{r crosstab-compact, fig.cap="**Table S4.** Compact cross-tabulation of all Vel native labels (rows) against PsychAD subclass assignment (columns EN%, IN%, Glia%). Vel EN-layer labels (L2-3, L4, L5, L6, L5-6-IT) map 93–100% to EN subtypes; canonical IN labels (SST, PV, VIP, CALB2, CCK, RELN, SV2C, SST_RELN, NOS, PV_MP) map 86–100% to IN subtypes. 'Interneurons' maps 96.1% EN despite its name. 'Progenitors' (43.3% EN / 25.3% IN / 25.6% Endo) and 'Glial_progenitors' (29.7% EN / 70% Glia) represent heterogeneous transitional populations."}
crosstab_compact <- tribble(
  ~`Vel native label`,           ~N,      ~`EN%`,  ~`IN%`,  ~`Glia%`, ~Interpretation,
  "Interneurons",                16411,    96.1,     1.6,     2.3,     "Immature upper-layer EN progenitors",
  "Protoplasmic_astrocytes",      4610,     0.5,     0.0,    99.5,     "Astrocyte",
  "INT",                          2922,     3.6,    88.2,     8.1,     "Genuine cortical interneurons",
  "OPC",                          2585,     0.2,     0.0,    99.8,     "OPC",
  "Glial_progenitors",            1450,    29.7,     0.3,    70.0,     "Mixed EN/glia transitional",
  "L4",                           1275,    99.6,     0.1,     0.3,     "Layer IV EN",
  "Progenitors",                  1271,    43.3,    26.6,    30.1,     "Mixed transitional (+ Endo)",
  "L2-3",                         1239,   100.0,     0.0,     0.0,     "Upper-layer EN (L2–3)",
  "Microglia",                    1075,     1.5,    95.5,     3.0,     "Microglia (Vel label maps to Micro)",
  "Oligos",                       1069,     0.0,     0.0,   100.0,     "Oligodendrocyte",
  "CALB2",                        1015,     6.9,    89.3,     3.8,     "Genuine IN",
  "PV",                            916,     9.4,    86.6,     4.0,     "Genuine PV IN",
  "SST",                           830,     2.0,    97.7,     0.2,     "Genuine SST IN",
  "L6",                            503,    97.4,     0.0,     2.6,     "Layer VI EN",
  "SST_RELN",                      488,     0.2,    99.8,     0.0,     "Genuine SST/RELN IN",
  "Fibrous_astrocytes",            460,     0.2,     0.2,    99.6,     "Astrocyte",
  "VIP",                           451,     1.1,    98.9,     0.0,     "Genuine VIP IN",
  "CCK",                           341,     0.0,   100.0,     0.0,     "Genuine IN",
  "RELN",                          238,     0.0,   100.0,     0.0,     "Genuine IN",
  "SV2C",                          226,     0.0,   100.0,     0.0,     "Genuine IN",
  "L5",                            183,    93.4,     0.0,     6.6,     "Layer V EN",
  "NOS",                           170,     3.5,    95.9,     0.6,     "Genuine IN",
  "PV_MP",                         152,     2.6,    95.4,     2.0,     "Genuine IN",
  "L5-6-IT",                       106,   100.0,     0.0,     0.0,     "Deep-layer IT EN",
  "SP",                             14,   100.0,     0.0,     0.0,     "Subplate EN"
)

knitr::kable(
  crosstab_compact,
  caption = "**Table S4.** Vel native label × PsychAD subclass cross-assignment (Experiment 2). All 40,000 Vel cells; sorted by N descending. EN% = fraction assigned to EN_ subtypes; IN% = fraction assigned to IN_ subtypes; Glia% = remainder. Note: 'Microglia' in Vel maps 95.5% to Micro/Microglia in PsychAD, which the diagnostic script classifies as IN-like — this reflects the Micro/Microglia label convention, not genuine inhibitory identity.",
  align = c("l","r","r","r","r","l"),
  format.args = list(big.mark = ",")
)
```

### Cell Composition Comparison by Age Bin

```{r composition-comparison, fig.height=5, fig.cap="**Figure S11.** Cell type composition (EN%, IN%, OPC%, Micro%) for Velmeshev and PsychAD cells in the <1y and 1–5y age bins, using PsychAD native subclass labels throughout (Experiment 2: Vel cells classified by scANVI; PsychAD cells = native annotation). At <1y, the two datasets are dramatically divergent: Vel is EN-dominated (61.7%) with low IN (17.9%), while PsychAD is IN-dominated (46.4%) with very low EN (6.7%). By 1–5y, IN% converges (Vel 20.5% vs PsychAD 22.7%), though PsychAD remains lower in EN (17.3% vs Vel 40.2%). The convergence of IN% by 1–5y is consistent with both datasets capturing the same maturing interneuron population; the persistent EN% discrepancy suggests PsychAD continues to undersample upper-layer excitatory neurons relative to Vel."}
composition_data <- tribble(
  ~age_bin, ~source,    ~EN_pct, ~IN_pct, ~OPC_pct, ~Micro_pct, ~n,
  "<1y",    "Vel",       61.7,    17.9,    5.3,       1.2,       20000,
  "<1y",    "PsychAD",    6.7,    46.4,   27.8,      12.2,       20000,
  "1–5y",   "Vel",       40.2,    20.5,   10.9,       4.2,       20000,
  "1–5y",   "PsychAD",   17.3,    22.7,   18.1,      10.7,       20000
) |>
  pivot_longer(c(EN_pct, IN_pct, OPC_pct, Micro_pct),
               names_to = "cell_class", values_to = "pct") |>
  mutate(
    cell_class = recode(cell_class,
      "EN_pct" = "EN", "IN_pct" = "IN", "OPC_pct" = "OPC", "Micro_pct" = "Microglia"
    ),
    cell_class = factor(cell_class, levels = c("EN","IN","OPC","Microglia")),
    source = factor(source, levels = c("Vel","PsychAD")),
    age_bin = factor(age_bin, levels = c("<1y","1–5y"))
  )

CELL_COLS <- c("EN" = "#2980B9", "IN" = "#E67E22", "OPC" = "#27AE60", "Microglia" = "#8E44AD")

ggplot(composition_data, aes(x = source, y = pct, fill = cell_class)) +
  geom_col(position = "dodge", width = 0.75, colour = "white") +
  facet_wrap(~age_bin, labeller = labeller(age_bin = c("<1y" = "Age < 1 year", "1–5y" = "Age 1–5 years"))) +
  scale_fill_manual(values = CELL_COLS) +
  scale_y_continuous(labels = label_percent(scale = 1),
                     expand = expansion(mult = c(0, 0.08))) +
  labs(
    x = NULL, y = "% of cells", fill = "Cell class",
    title = "Cell composition: Vel vs PsychAD by age bin",
    subtitle = "Vel EN% from scANVI under PsychAD supervision; PsychAD % from native HBCC labels"
  ) +
  theme_pub() +
  theme(legend.position = "right")
```

```{r en-in-ratio-table}
en_in_summary <- tribble(
  ~`Age bin`, ~Source,    ~N,      ~`EN%`, ~`IN%`, ~`EN/IN ratio`, ~`OPC%`, ~`Micro%`,
  "<1y",      "Vel",      20000,   61.7,   17.9,   3.45,            5.3,     1.2,
  "<1y",      "PsychAD",  20000,    6.7,   46.4,   0.14,           27.8,    12.2,
  "1–5y",     "Vel",      20000,   40.2,   20.5,   1.96,           10.9,     4.2,
  "1–5y",     "PsychAD",  20000,   17.3,   22.7,   0.76,           18.1,    10.7
)

knitr::kable(
  en_in_summary,
  caption = "**Table S5.** EN/IN/OPC/Microglia fractions for Vel and PsychAD cells by age bin. Vel fractions are from scANVI classification under PsychAD supervision (Experiment 2); PsychAD fractions are from native HBCC subclass labels. The EN/IN ratio inverts between datasets at <1y (Vel 3.45 vs PsychAD 0.14). By 1–5y, IN% converges to within 2 percentage points (Vel 20.5% vs PsychAD 22.7%), while EN% remains 2.3× higher in Vel (40.2% vs 17.3%).",
  align = c("l","l","r","r","r","r","r","r"),
  format.args = list(big.mark = ",")
)
```

---

## Brain Region Confirmation

A prerequisite for attributing the EN/IN composition difference to biology rather than tissue
sampling is that both datasets sample the same brain region. This section documents the brain
region filtering applied and identifies a residual ambiguity in tissue granularity.

### Filter Applied: `pfc_only: true`

Both VELMESHEV and PSYCHAD sources were processed with `pfc_only: true` in the pipeline
configuration, which retains only cells where the harmonised `region` metadata equals
`"prefrontal cortex"`. After this filter:

```{r brain-region-table}
region_summary <- tribble(
  ~Source,    ~`N cells`, ~`region (harmonised)`, ~`tissue (raw metadata)`,
  "VELMESHEV", 40000,     "prefrontal cortex",     "frontal cortex",
  "PSYCHAD",   42408,     "prefrontal cortex",     "dorsolateral prefrontal cortex"
)

knitr::kable(
  region_summary,
  caption = "**Table S6.** Brain region metadata for cells retained in the VelPsychAD cross-assignment experiments (post pfc_only filter). The harmonised `region` field is identical for both sources. The raw `tissue` metadata reveals a specificity difference: PsychAD cells carry an anatomically specific label (dlPFC) while Velmeshev cells are labelled at the coarser 'frontal cortex' level.",
  align = c("l","r","l","l"),
  format.args = list(big.mark = ",")
)
```

Both sources show 100% `region == "prefrontal cortex"` after filtering. Brain region has
therefore been successfully harmonised at the PFC level — it is not an uncontrolled confound.

### Residual Concern: Tissue Granularity

The raw `tissue` metadata reveals that the two sources differ in anatomical specificity:

- **PsychAD (HBCC cohort):** `tissue = "dorsolateral prefrontal cortex"` — a specific
  subregion of PFC (Brodmann areas 9/46), sampled by the HBCC postmortem dissection protocol.
- **Velmeshev:** `tissue = "frontal cortex"` — a coarser label. The original Velmeshev
  dataset spans multiple frontal regions including BA9, BA8, BA46, and frontoparietal cortex;
  the `pfc_only` filter retains only cells annotated as "frontal cortex" in the CellxGene
  harmonised metadata, but this includes any gyral/sulcal variation within that label.

**Implication:** Both datasets sample the prefrontal cortex, and the high-level brain region
is not a confound. However, the Vel "frontal cortex" label may include cells from PFC
subregions not sampled by the HBCC dlPFC dissection. Given that cell type composition
(particularly EN/IN balance across cortical layers) can vary across PFC subregions
within the same developmental window, a residual laminar or subregional confound cannot be
excluded. This is unlikely to explain the 9-fold difference in EN/IN ratio at <1y
(Vel 3.45 vs PsychAD 0.14), but could contribute a fraction of the remaining discrepancy.

---

## Synthesis and Conclusions

### The Vel "Interneurons" Label Resolved

**Velmeshev "Interneurons" are excitatory progenitors, not inhibitory neurons.**
Under PsychAD supervision (Experiment 2), 96.1% of Vel "Interneurons" are classified as
EN subtypes — primarily EN_L2_3_IT (45.6%, confidence 0.94), a fate-committed upper-layer
excitatory type. This is consistent with the high SATB2 expression documented in earlier
sections of this report. By contrast, Vel "INT" cells map 88.2% to canonical inhibitory
types (PV, SST, VIP, LAMP5) — confirming that Vel uses two distinct IN-like labels with
opposite transcriptional identities.

The developmental context explains the naming convention: at birth, the upper cortical
layers are populated by post-mitotic excitatory neurons that are morphologically
undifferentiated and co-express some transcription factors shared with early inhibitory
lineages. Velmeshev likely applied the "Interneurons" label based on morphology or
intermediate marker profiles rather than strict GAD1/GAD2 positivity.

### The PsychAD <1y IN-Enrichment Is Genuine Biology

**PsychAD <1y dlPFC is genuinely enriched for cortical interneurons (46.4% IN) relative to
Velmeshev frontal cortex (17.9% IN).** Three independent lines of evidence converge on this
conclusion:

1. **Raw marker expression** (Sections 3–4): PsychAD <1y shows 4–11× lower EN marker CPM
   and 4–5× lower EN frac_nonzero than Vel, at comparable library depth.
2. **Experiment 1** (Vel labels → classify PsychAD): PsychAD <1y cells land on SST, PV, VIP
   and other canonical IN types (46.7% total IN-type), not on Vel "Interneurons" (2.4%).
3. **Experiment 2** (PsychAD labels → classify Vel): PsychAD <1y native label distribution
   shows 46.4% IN subtypes (SST 13.4%, VIP 12.5%, PV 10.0%) — confirmed by the HBCC cohort's
   own annotation pipeline independently of scANVI.

### Biological Interpretation

The composition difference between Vel and PsychAD at <1y is most likely explained by a
combination of:

| Factor | Evidence | Contribution |
|--------|----------|--------------|
| **Genuine EN/IN developmental difference** | PsychAD EN% matches HBCC native annotation (6.6%); persistent across all 11 donors; no age gradient within cohort | Primary driver; ~4–5× EN deficit |
| **Dissection depth (laminar sampling)** | PsychAD = specific dlPFC (BA9/46); Vel = broader frontal cortex. EN/IN balance varies by cortical layer and developmental stage | Secondary driver; hard to quantify |
| **Cohort biology** | PsychAD HBCC = postmortem cohort; Vel = autism+control cohort (UCSF), dissection and processing may differ | Cannot be excluded |
| **Annotation bias (scANVI, corrected in v4a)** | Original 5% EN → 6.7% after removing adult supervision; confirms ~2× scANVI suppression beyond the biological signal | Modest (~2×), corrected |

: **Table S7.** Factors contributing to the EN% difference between PsychAD and Vel at <1y. {tbl-colwidths="[22,53,25]"}

### Convergence at 1–5 Years

By the 1–5y window, IN% converges between the two datasets (Vel 20.5% vs PsychAD 22.7%,
a difference of only 2.2 pp). This developmental convergence is consistent with the
biological interpretation: the rapid EN progenitor expansion phase in Vel <1y cells
(dominated by "Interneurons"/EN_L2_3_IT progenitors) is complete by 1 year, and both
datasets then sample a maturing cortex with similar IN proportions. The continued EN%
discrepancy at 1–5y (Vel 40.2% vs PsychAD 17.3%) likely reflects the persistent laminar
or cohort differences noted above.

### Implications for GRN Analysis

For downstream gene regulatory network (GRN) inference using pseudobulk RNA-seq:

1. **Cell type labels should use the corrected v4a scANVI assignments** (PsychAD <1y EN% = 6.7%,
   reflecting the biological reality of low EN content) rather than the uncorrected v3 assignments.

2. **PsychAD and Vel EN populations are not equivalent at <1y.** Vel EN includes a large
   proportion of EN_L2_3_IT progenitors (45.6% of the "Interneurons" class); PsychAD EN at
   <1y is dominated by deep-layer types (EN_L6B 2.3%, EN_L6_CT 1.7%). GRN comparisons
   across datasets in the EN class should stratify by EN subtype, not treat EN as monolithic.

3. **PsychAD <1y IN cells are reliable.** The cardinal IN subtypes (SST, PV, VIP) in PsychAD
   <1y are consistently identified across all three classification approaches (native annotation,
   Vel-supervised scANVI, PsychAD-supervised cross-classification). IN GRN analysis in the
   PsychAD <1y window can proceed with confidence.

4. **The 1–5y window is the most comparable.** Both datasets converge in IN% by 1–5y;
   cross-dataset GRN comparisons are most reliable in this age window.

---

## Session Information {.unnumbered}

```{r session-info}
#| code-fold: false
sessionInfo()
```
